---
title: "Final_Exam"
author: "Keith Lee"
date: '2022-10-26'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dslabs)
library(ggplot2)
library(tidyverse)
```
# 1. 
**(a). How many observations and variable are there? What are the types of each variable?**
```{r}
full_data <- us_contagious_diseases
dim(full_data) #to find number of variables and observations
str(full_data) #str() shows the types of each variable
```
Disease and state are factor variables which is also another term for categorical variables. Weeks_report, year, count, and population are numeric variables. However, year can also be considered as categorical variable.  

**(b). Compute the frequency of each type of disease and visualize the proportion using barchart.**
```{r}
table(full_data$disease) #table() shows how many observations exist in each variable
ggplot(data=full_data) + geom_bar(mapping=aes(x=disease)) #used bar chart to easily observe the overall difference of each diseases. However, it is not useful to see specific value of each observation. 
```

**(c). Compute the 0.1, 0.5, 0.9 quantiles of the population for each type of diseases. Write a paragraph to compare the quantiles of the population across the disease.**
```{r}
full_data %>%
  group_by(disease) %>%
  summarize(quantile_0.1 = quantile(count, probs=0.1, na.rm=TRUE))

full_data %>%
  group_by(disease) %>%
  summarize(quantile_0.5 = quantile(count, probs=0.5, na.rm=TRUE))

full_data %>%
  group_by(disease) %>%
  summarize(quantile_0.9 = quantile(count, probs=0.9, na.rm=TRUE))
```
For 10% of counts, we had proportion of 10 cases of Hepatitis A, 4 cases of Pertussis, and the rest 0. For 90% of counts, we had proportion of 1013.5 cases of Hepatitis A, 13697 cases for Measles, 1163 cases for Mumps, 1953 cases for Pertussis, 634 cases for Polio, 631.4 cases for Rubella, and 537.6 cases for Smallpox. For median values, we have 138.5 for Hepatitis A, 577 for Measles, 29 for Mumps, 81 for Pertussis, 57 for Polio, 5 for Rubella, and 8 for Smallpox. 

# 2. 
**Find the top 5 states with the most "Mumps" cases over the 10 years from 1991 to 2000 (both years inclusive). Find the bottom 5 states with the least Hepatitis A cases over the 5 years from 1994 to 1998 (both years inclusive).**
Most
```{r}
full_data %>% #used pipe operator for chaining commands
  filter(year >= 1991, year <= 2000, disease == "Mumps") %>% #filer() for getting the range of year and specific disease we want
  arrange(desc(count)) %>% #arrange() gives in order from lowest to highest, but also used desc() to have descending order, from highest to lowest
  head(5) #used head() function to get only first 5 observations
```
Least
```{r}
full_data %>% #used pipe operator for chaining commands
  filter(year >= 1994, year <= 1998, disease == "Hepatitis A") %>% #filer() for getting the range of year and specific disease we want
  arrange((count)) %>% #arrange() gives in order from lowest to highest
  head(5) #used head() function to get only first 5 observations
```
# 3. 
**For the state of Texas, (a) Add a variable ave_count, representing the average reported case count per weeks_reporting for each year.**
```{r}
full_data %>%
  filter(state == "Texas") %>% #filter to get only state of Texas
  group_by(year) %>% #grouped by each year
  mutate(ave_count = count/weeks_reporting) #added ave_count variable by mutate()
```

**(b). Plot ave_count against the year while using different colors for different diseases.**
```{r}
full_data %>%
  filter(state == "Texas") %>% #used filter to get state of Texas
  group_by(year) %>% #grouped by each year
  mutate(ave_count = count/weeks_reporting) %>% #add ave_count variable to the data
  ggplot(mapping = aes(x=year, y=ave_count, color=disease)) + geom_line() #ggplot + geom_line to plot year against ave_count
```

**(c). Remove all the observations for disease "Measles" and redo the plot in (b).**
```{r}
full_data %>%
  subset(disease != "Measles") %>% #removed Measles by using subset() and !=
  filter(state == "Texas") %>% #to get state of Texas
  group_by(year) %>% #grouped by each year
  mutate(ave_count = count/weeks_reporting) %>% #add ave_count variable to the data
  ggplot(mapping = aes(x=year, y=ave_count, color=disease)) + geom_line() #line graph to see year against ave_count
```

# 4. 
**Redo the problem 3 for the state of New York. Write a paragraph to compare the results of the two states.**
**(a)**
```{r}
full_data %>%
  filter(state == "New York") %>%
  mutate(ave_count = count/weeks_reporting) %>%
  head(20)
```

**(b)**
```{r}
full_data %>%
  filter(state == "New York") %>%
  mutate(ave_count = count/weeks_reporting) %>%
  ggplot(mapping = aes(x=year, y=ave_count, color=disease)) + geom_line()
```

**(c)**
```{r}
full_data %>%
  subset(disease != "Measles") %>%
  filter(state == "New York") %>%
  mutate(ave_count = count/weeks_reporting) %>%
  ggplot(mapping = aes(x=year, y=ave_count, color=disease)) + geom_line()
```
From the line graph that we created previously for Texas and New York, we can first clearly see that Measles had far more counts than other diseases. Also, measles prevalence peaked in the 1940s in New York and 1950s in Texas. After removing the measles, we can see that pertussis had most prevalence rate in the 1940s in New York. Also, in Texas, pertussis had the most counts from the 1940s to 1960s and started to decrease after the 1960s. Polio had more cases in New York than in Texas from the 1930s to the 1960s. We had very few counts for smallpox in both New York and Texas. For rubella, Texas peaked in the 1970s, while New York had relatively low cases. New York had Hepatitis cases in the mid-1960s, peaked in 1970, then decreased afterward. Texas also had a Hepatitis case in the 1960s, but it didn't decrease and kept the pace until 2000. We had few instances of mumps in New York, but Texas had relatively high cases. 

# 5. 
**(a)** 
**For each state and year, find the total count of all diseases for the given state and year.**
```{r}
full_data %>%
  group_by(year, state) %>% #groupd by each year and state
  summarise(total_count = sum(count)) #used to summarise to create a new data frame which gives total count of each year and state
```

**(b)**
**For each state and year, find the disease count density, which is defined by the total count of all diseases divided by the population for the given state and year.**
```{r}
full_data %>%
  group_by(year, state) %>% #grouped by each year and state
  summarize(disease_count_density = sum(count)/mean(population)) #used summarise to create a disease_count_density data frame and diveded the sum of count by mean of population to get the density
```

# 6. 
**(a). Find the 3 state and year pairs that have the largest total count of all diseases.**
```{r}
full_data %>%
  group_by(state, year) %>% #to get pair of each year and state
  summarize(total_count = sum(count)) %>% #to make total count of all disease in each year and state
  arrange(desc(total_count)) #arranged hihgest to lowest to get the largest count
```
Pennsylvania 1938, New York 1941, and Pennsylvania 1941 had the largest count of all diseases.

**(b). Fint the 3 state and year paris that have the largest disease count density.**
```{r}
full_data %>%
  group_by(state, year) %>% #paired each year and state
  summarize(disease_count_density = sum(count)/mean(population)) %>% #to get count density of each count divided by population mean 
  arrange(desc(disease_count_density))  #arranged highest to lowest to get all diseases
```
Vermont 1936, Utah 1942, Wisconsin 1938 had the largest disease count density. 